Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-12 doc/manuscript_hkk_drought_sensitivity.Rmd
edits to discussion
@krishnaanujan , I've now been through the entire manuscript. I may provide some more edits later, but please proceed with passing along to other coauthors

<!----Limitations in filling gap/ comparing only two droughts---->
While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size. 
<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments. 
Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.  -->
<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig. `r fig_droughts`).
Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
While the meteorological data justify our characterization of these droughts as "dry-season" (2010) and "wet-season" (2015), growth records from numerous additional droughts would be necessary to rigorously generalize differences based on the seasonality of drought.
Given the relative rarity of drought and the lack of multi-decadal records of annual tree growth for diverse tropical forests, there is currently no data set on which such an analysis could be performed.
In temperate and boreal forests, where the majority of species form annual growth rings, tree-ring analysis can disentangle species and individual sensitivities to larger numbers of droughts (although sample sizes still remain limited).
Studies in temperate forests have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
As tropical dendrometer band records continues to grow, so will the potential to better understand how drought characteristics interact with species traits and microenironment to shape growth respones.

@@

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts. 
An extremely important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community -- particularly because ring formation is caused by dry-season dormancy.
Our finding that a wet-season drought had greater impact than a dry-season drought calls into question the generality of previous findings based on tree-ring chronologies from HKK  [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation.
Care must be taken in interpretation, however, as the 2015 drought was stronger (Fig 1).
<!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.  -->
*(KAT paused here.)*
*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S5). 
More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. 
In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences. 
Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought in alignment with recent global analyses (@liu_drought_2022).

<!----Science is better now---->
Drought sensitivity is often defined as a static species trait but our study reveals expanded dimensions of drought responses with possible consequences for management and restoration. 
Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation. 
Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
<!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
More inclusive frameworks for drought responses like @trugman_why_2021 (on mortality) encompass ecological complexity and covarying traits with environment like, but they assume consistent sensitivity of individual trees across events.
Our results, however, demonstrate an additional temporal dimension of variation, at least partially attributable to drought characteristics. 
Our results indicate that drought responses in tropical trees may not be a fixed response across events, but jointly shaped by species adaptations and local environmental conditions. 
Given these inconsistencies, diverse stands that may represent greater variation in ecological and evolutionary strategies, may have mechanisms for stable functioning under future droughts.

>>>>

<!----Limitations in filling gap/ comparing only two droughts---->
While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size. 
<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments. 
Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.  -->
<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig. `r fig_droughts`).
Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
While our finding that a wet-season drought (2015) had greater impact on woody growth than a dry-season drought (2010) calls into question the generality of previous findings based on tree-ring chronologies from HKK [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation, care must be taken in interpretation given that the 2015 drought was more anomalous on multi-month time scales (Fig. `r fig_droughts`).
Rigorous generalization on tropical tree growth responses to any specific drought characteristic would require annual growth records from numerous species over numerous droughts, but no such data set currently exists.
Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
As tropical dendrometer band records continues to grow, so will the potential to better understand how specific drought characteristics interact with species traits and microenironment to shape growth respones.

@@

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig. S5).
More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to directly and interactively influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).

<!----Science is better now---->
Our study reveals that drought sensitivity is not a static trait associated to particular species, traits, or microenvironments, but rather a dynamic response shaped interactively by these factors and drought characteristics. 
Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation. 
Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
<!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
More inclusive frameworks for drought responses encompass ecological complexity and covarying traits with environment [e.g., @trugman_why_2021] but still assume consistent sensitivity of individual trees across events.
However, we demonstrate that drought responses of tropical trees are not linearly predictable across drought events, but rather are shaped by the unique characteristics of each drought based on interactions between species adaptations and local environmental conditions.
Without overriding the net negative effects of drought on tropical tree growth, these complexities reduce the likelihood of any given species or tree suffering severe growth declines under multiple consecutive droughts, thereby promoting diversity and stabilizing functioning of diverse tropical forests under drought.
